---
title: 'Activity: Statistical Data Manipulation'
author: "Reister L. RUedas"
date: "5/14/2021"
output: html_document
---

```{r}
library(ggplot2)
library(plyr)
library(reshape2)
library(knitr)
```

```{r}
data("mtcars")
mtcars
```

Mean - Median - Mode

Mean: Calculated by taking the sum of the values and dividing with the number of values.

Median: Middle most value. The median() function is used in R to calculate.

Mode: Highest number of occurences in a set of data. Unlike mean and median, mode can have both numeric and character data.

```{r}
#Getting the mean of members in the dataset anime
mean(mtcars$hp)
#Getting the median of members in the dataset anime
median(mtcars$hp)
#Getting the mode of members in the dataset anime
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
getmode(mtcars$hp)
```
Max - Min - Range

Max: Largest value in the dataset

Min: Smallest value in the dataset

Range: Interval difference between the lowest and the highest value

```{r}
#Getting the max of members in the dataset anime
max(animes$members)
#Getting the min of members in the dataset anime
min(animes$members)
#Getting the range of members in the dataset anime
range(animes$members)
```
Standard Deviation

  An observation variable that is the square root of its variance.

```{r}
sd(animes$members)
```
P-Value
  Measures the probability of an observed difference that could have occurred just by random chance.

```{r}
z <- 12
x <- 14
c <- 05
xbar <- 2
2*(1 - pnorm(xbar, mean = z, sd = x/sqrt(20)))
```
Variance
  Refers to a statistical measurement of the spread between numbers in a dataset. Variance also measures how far each number in the set is from the mean and from every other number in the set.

```{r}
var(animes$members)
```
Residual
  Simple linear regression model is the difference between the observed data of the dependent variable y and the fitted values ŷ.

```{r}
newdata <- lm(mpg ~ wt + cyl, data = mtcars)
resid(newdata)
```