---
title: Exercise 3
author: Reister L. Ruedas
output: html_document
---

**Note:** We will assume that the data file `expenditures_lec2.csv` is in your working directory.

Problem 1: Load the CSV file `expenditures_lec2.csv` into memory, using the command `read.file()` or `read.csv()` as described in lecture. Save the result as a variable.

```{r}
# Fill in, with comments added:
expenditures <- read.csv(file = 'expenditures_lec2.csv')
expenditures
library(plyr)
library(ggplot2)
```

Problem 2: Create a scatter plot showing how `healthcare` costs behave as a function of the `age.respondent` variable. Don't forget to load any libraries you need!

```{r fig.width=6, fig.height=3, dpi=200, fig.align='center', cache=TRUE}
# Fill in, with comments added:
ggplot(data = expenditures, mapping = aes(x = healthcare, y = age.respondent)) + geom_point()
```

Problem 3: Create a bar chart showing how many rows in the data set are in each `region`. 

```{r fig.width=6, fig.height=4, dpi=50, fig.align='center', cache=TRUE}
# Fill in, with comments added:
ggplot(expenditures, aes(x = region)) + geom_bar()

```

Problem 4: Use functions from lecture 2 (such as the `plyr` package) to compute the fraction (or percentage) of households that are in each `region`, instead of the raw counts. Then plot this as a bar chart.

```{r fig.width=6, fig.height=3, dpi=100, fig.align='center'}
# Fill in, with comments added:
total = ddply(expenditures, "region", summarize, households = sum(housing))
percentage = ddply(total, "region", summarize, percent = households/ sum(expenditures$housing))

ggplot(percentage, aes(x = region, y = percent)) + geom_bar(stat = "identity")

```

Problem 5: Remake the plot of problem 2, showing `healthcare` costs, but this time color-code the points to distinguish between those which spent over and under $3000 on health care.

```{r fig.width=6, fig.height=3, dpi=100, fig.align='center'}
# Fill in, with comments added:
ggplot(data = expenditures, mapping = aes(x = healthcare, y = age.respondent, color = healthcare)) + geom_point()

```
