---
title: "Mini Project: Data Exploration"
author: "Humphrey Demoulder Pimentel, Nathaniel Cayco, & Reister Ruedas"
date: "6/8/2021"
output: html_document
---


```{r setup, include=FALSE}
### Load Libary
library(ggplot2)
library(plyr)
library(reshape2)

### Load Data
data(nlschools, package = 'MASS')
```

# Variable Description
#### This data frame contains 2287 rows and the following columns:
### lang

Language test score.

### IQ

Verbal IQ.

### class

Class ID.

### GS

Class size: number of eighth-grade pupils recorded in the class (there may be others: see COMB, and some may have been omitted with missing values).

### SES

Social-economic status of pupil's family.

### COMB

Were the pupils taught in a multi-grade class (0/1)? Classes which contained pupils from grades 7 and 8 are coded 1, but only eighth-graders were tested.



```{r cars}
# Explore Data set at show basic informations
head(nlschools)
# Show summary of the data frame
summary(nlschools)
# Show data type of variables in the data fram nlschools
str(nlschools)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
