---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
<style>
 td {
  width: 150px;
}
</style>
---
title: Spot Check Quiz No. 1
author: Reister L. Ruedas
output: html_document
---

#### Instructions 

The due date is 11:59pm May 29 2019. Submit this Rmd file on blackboard. Don't submit additional files.

Fill in the missing code chunks to answer the following questions. You should use a reasonable coding style and comment your code so that it is clear what you are doing. As a rough rule of thumb, expect to add at least one comment to each missing code chunk.

If the question asks for a numerical answer, such as "what fraction of households in the data set satisfy condition X," do not hard code the value from the computation. If the data set changes, your answer should  update automatically. 

Here are two recommended ways that you can report your answer:

**Example Question 1:** In the `cars` data set, what is the average value of the `speed` column?

```{r}
# take average of mpg$speed, save as variable
avg.speed = mean(cars$speed)
```

**Answer:** The average value was `r avg.speed`

**Example Question 1:** In the `ChickWeight` dataset, what was the average weight of the chickens grouped by each diet type?

```{r}
library(plyr)
library(knitr) # for the kable() function, which prints data frames as tables:

# group the chickens by diet and take average weight
avg.weights = ddply(ChickWeight, "Diet", summarize, avg.weight = mean(weight))
# display the data frame as an easy-to-read table, with 3 significant digits
kable(avg.weights, digits = 3, col.names = c('Diet', 'Average Weight'))
```


#### Questions 

**Note:** We will assume that the data file `dataset_hw1.csv` is in your working directory.

**Problem 1:** 

Load the CSV file `dataset_hw1.csv` into memory, using the command `read.file()` or `read.csv()` as described in lecture. Save the result as a variable called `expenditures`. 

Hint: You may want to look at the CSV file using another program such as excel to see if the first row is a header. 

```{r}
# Fill in, with comments added:
# Load dataset as expenditures
expenditures <- read.csv(file = 'dataset_hw1.csv', stringsAsFactors=T)
dim(expenditures)

# 
```

**Problem 2:**

a) How many variables are in the data set? 

b) The column `race.ref.person` describes the race of each respondent. What are the different levels that this column is allowed to take?
```{r}
# Fill in, with comments added:
# Count the number of columns on the data set and print it
col = ncol(expenditures)
col
# Get the levels for column of race.ref.person
levels(expenditures$race.ref.person)
```

**ANSWER:** 

a) 16

b) "asian","black","multi-race","native american","pacific islander","white"

**Problem 3:**

What fraction of records in the data set belong to each `race.ref.person`?

```{r}
# Fill in, with comments added:
# Compute for each data set of race.ref.person
counts = ddply(expenditures, c("race.ref.person"), summarize, n.households = length(housing))
counts 
# Get the percentage of the data set 
ddply(counts, "race.ref.person", mutate, percent = n.households / sum(counts$n.households))

```

**Problem 4:**

Change the `expenditures` data frame so that `race.ref.person` column has only 3 coded types: 

1. black
2. white
3. others (all else)

Print the number of records in each `race.ref.person` as a table.

```{r}
# Fill in, with comments added:
# Map the values from levels to the required 3 coded types
expenditures$race.ref.person = mapvalues(expenditures$race.ref.person,
                               from = levels(expenditures$race.ref.person),
                               to = c('others', 'black', 'others', 'others', 'others', 'white'))
# Display the data frame as an easy-to-read table
counts = ddply(expenditures, c("race.ref.person"), summarize, n.records = length(race.ref.person))
kable(counts, col.names = c("race.ref.person", "n.records"))     

```

**Problem 5:**


What fraction of the general population live in the Midwest, Northeast, South, and West regions of the US? What fraction of those in the 96% income percentile (i.e., the "96%") live in these regions? Do the percentages differ for the general population and the "96%"?

Create a data frame with the following columns:

1. `region`: Midwest, Northeast, South, West, and NA (not everyone reports their region!)
2. `general.pop`: the percentage of households in `expenditures` in each region
3. `rich.pop`: the percentage of households whose `income.rank` is greater than 0.96 in each region

(as a sanity check, both the `general.pop` and `rich.pop` columns should sum to 1)

Print this data frame as a table. Hint: you may want to use the function `merge()` to combine two data frames. (see the self-study material for lecture 2)

What are the main differences in the percentages for the general and rich population?

```{r echo= "true"}
# Fill in, with comments added:
# Data Frame for region
region.df = data.frame(region = levels(expenditures$region))
region.df
# Data Frame for the percentage of households en each region
general.pop = ddply(expenditures, 'region', summarize, general.pop = sum(length(housing)) / sum(length(expenditures$housing)))
# Merge Region data frame and general population data frame
merged.df = merge(x =general.pop, y = region.df , by ="region", all.x = TRUE)
merged.df
# Data Frame for the rich population. Greater than 0.96 in each region. Computation 
rich.pop = ddply(expenditures, 'region', summarize, rich.pop = sum(length(which(income.rank > 0.96)), na.rm = TRUE)/ sum(length(expenditures$income.rank > 0.96), na.rm = TRUE))
rich.pop
# Merge again 
merged.df = merge(x = merged.df, 'region', y = rich.pop, by = 'region', all.x = TRUE)
merged.df

sum(merged.df$general.pop)
sum(merged.df$rich.pop)

# The Main difference on the percentages is it may show different result. The total percentage for the rich population show it's not equal to 1 because there are factors that affect this value like the problem only considered greater than 0.96. The problem don't expect the rich population to reach 100 percent since it didn't include other values.

```
