<style>
 td {
  width: 150px;
}
</style>
---
title: 95-868 Homework 1
author: YOUR NAME HERE
output: html_document
---

#### Instructions 

The due date is 11:59pm May 29 2019. Submit this Rmd file on blackboard. Don't submit additional files.

Fill in the missing code chunks to answer the following questions. You should use a reasonable coding style and comment your code so that it is clear what you are doing. As a rough rule of thumb, expect to add at least one comment to each missing code chunk.

If the question asks for a numerical answer, such as "what fraction of households in the data set satisfy condition X," do not hard code the value from the computation. If the data set changes, your answer should  update automatically. 

Here are two recommended ways that you can report your answer:

**Example Question 1:** In the `cars` data set, what is the average value of the `speed` column?

```{r}
# take average of mpg$speed, save as variable
avg.speed = mean(cars$speed)
```

**Answer:** The average value was `r avg.speed`

**Example Question 1:** In the `ChickWeight` dataset, what was the average weight of the chickens grouped by each diet type?

```{r}
library(plyr)
library(knitr) # for the kable() function, which prints data frames as tables:

# group the chickens by diet and take average weight
avg.weights = ddply(ChickWeight, "Diet", summarize, avg.weight = mean(weight))
# display the data frame as an easy-to-read table, with 3 significant digits
kable(avg.weights, digits = 3, col.names = c('Diet', 'Average Weight'))
```


#### Questions 

**Note:** We will assume that the data file `dataset_hw1.csv` is in your working directory.

**Problem 1:** 

Load the CSV file `dataset_hw1.csv` into memory, using the command `read.file()` or `read.csv()` as described in lecture. Save the result as a variable called `expenditures`. 

Hint: You may want to look at the CSV file using another program such as excel to see if the first row is a header. 

```{r}
# Loading 'dataset_hw1.csv' using read.csv(). By inspection, the first row contains the header.

expenditures = read.csv(file = 'dataset_hw1.csv', header = TRUE, sep = ",")
```

**Problem 2:**

a) How many variables are in the data set? 

b) The column `race.ref.person` describes the race of each respondent. What are the different levels that this column is allowed to take?
```{r}
# # Count the number of variables by counting the number of columns of the dataset using ncol() command.
NumVar = ncol(expenditures)
# List the different  levels of the race of each of the respondent.
Race <- levels(expenditures$race.ref.person)
```

**ANSWER:** 

a) There are `r NumVar` variables in the data set.

b) The allowable races are: `r Race`

**Problem 3:**

What fraction of records in the data set belong to each `race.ref.person`?

```{r}
# Count the number of the respondents with respect to their races.
library(plyr)
Race_Cnt = ddply(expenditures, "race.ref.person", summarize, Num.Race = length(race.ref.person))
#Compute for the fraction of the different races with respect to the total respondents
Race_Fraction = ddply(Race_Cnt, "race.ref.person", mutate, Race.Fraction = Num.Race/sum(Race_Cnt$Num.Race))
Race_Fraction
```

**Problem 4:**

Change the `expenditures` data frame so that `race.ref.person` column has only 3 coded types: 

1. black
2. white
3. others (all else)

Print the number of records in each `race.ref.person` as a table.

```{r}
# Change the all the race into others (all else) except for black and white
expenditures$race.ref.person = mapvalues(expenditures$race.ref.person, from = c("multi-race", "native american", "pacific islander", "white"), to = c("others(all else)", "others(all else)", "others(all else)", "others(all else)") )
#Print the number of respondents based on their races (black, white, and others)
NewRace = ddply (expenditures, "race.ref.person", summarize, NumRace = length(race.ref.person))
NewRace

```

**Problem 5:**


What fraction of the general population live in the Midwest, Northeast, South, and West regions of the US? What fraction of those in the 96% income percentile (i.e., the "96%") live in these regions? Do the percentages differ for the general population and the "96%"?

Create a data frame with the following columns:

1. `region`: Midwest, Northeast, South, West, and NA (not everyone reports their region!)
2. `general.pop`: the percentage of households in `expenditures` in each region
3. `rich.pop`: the percentage of households whose `income.rank` is greater than 0.96 in each region

(as a sanity check, both the `general.pop` and `rich.pop` columns should sum to 1)

Print this data frame as a table. Hint: you may want to use the function `merge()` to combine two data frames. (see the self-study material for lecture 2)

What are the main differences in the percentages for the general and rich population?



```{r}
# Create a household table indicating the number of records in each region
GenPop = ddply(expenditures, "region", summarize, GeneralPop = length(region))
# Create a list of regions where 'income.rank'is greater than 0.96
RichPop = subset (expenditures, "region", subset = income.rank > 0.96)

#Create a household table indicating the number of records of  the rich population where 'income,rank' is greater than 0.96, in each region.
CntRich = ddply(RichPop, "region", summarize, Rich_Pop= length(region))

#Create a table of the percentage of households in each region using the created GenPop table.
NewGenPop = ddply (GenPop, "region", mutate, general.pop = GeneralPop/ sum(GenPop$GeneralPop)*100)
NewGenPop
#Create a table of the percentage of households whose `income.rank` is greater than 0.96 in each region
NewRichPop = ddply (CntRich, "region", mutate, rich.pop= Rich_Pop/ sum(CntRich$Rich_Pop) *100)
NewRichPop


# Merge the the created table for the general (NewGenPop)  and rich (NewRichPop) population.
PopData = merge(x = NewGenPop, y = NewRichPop, by.x = "region", by.y = "region", all.x = TRUE)
PopData
# For 'select ()' function, to select a specific data frame and print it as a table.
library(dplyr)
# Select a specific data (column selection) to display.
NewPopData = select(PopData, c("region", "general.pop", "rich.pop"))
# Print the 'NewPopData' Table containing the following columns: 'region', 'general.pop' and 'rich.pop'
NewPopData

# Merge the the created table for the general (NewGenPop)  and rich (NewRichPop) population.
Diff_GenRichPop = ddply(NewPopData, "region", mutate, Diff_GR_Pop = general.pop - rich.pop)
# Print the 'Diff_GenRichPop' Table containing the main differences in the percentages for the general and rich population
Diff_GenRichPop



```

